---
title: "dplyr Notes"
author: "Christopher Willis"
output:
  html_document:
    theme: united
    highlight: tango
    df_print: paged
    toc: true
    toc_float: true
    collapsed: false
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = FALSE)
```

# The dplyr Package

```{r, echo=TRUE, results="hide"}
library(dplyr)
```

The `dplyr` package was built for the sole purpose of simplifying the process of manipulating, sorting, summarizing, and joining data frames. The basic `dplyr` functions (or "verbs") are:  
- select: select columns  
- filter: select rows  
- group by: group data into categorical levels  
- summarise: summarizes data by function of choice  
- arrange: reorder rows  
- join: join separate data frames  
- mutate: create features from existing features  
 
`dplyr` operations always take a data frame as their first argument; the result of such an operation is a transformed data frame. This allows operations to be chained together with the the pipe operator `%>%`.

# Quick View Operations

```{r}
glimpse(mtcars)
```

# Tibbles

The function `tibble` constructs a dplyr data frame, which allows for advanced printing, and some other features missing in base R data frames.
```{r}
x <- c(4, 7, 8, 2)
y <- 1:4
df <- tibble(x, y)
```

It is also possible to cast an `R` data frame as `dplyr` tibble
```{r}
df <- as_tibble(mtcars)
df
```

# Select

**Objective**: Reduce dataframe size to only desired variables for current task

**Function**: 
```
select(data, ...)
```

**Arguments**:
```
data:           data frame
...:            variable names or positions; useful expressions include -, :, and c().
```

**Examples**

select by column names
```{r}
mtcars.small <- mtcars %>% 
  select(mpg, cyl, wt)
```

select by column indices
```{r}
mtcars.small <- mtcars %>% 
  select(1:4)
```

select by column starting letter
```{r}
mtcars.small <- mtcars %>% 
  select(starts_with("c"))
```

**Scoped Variants**

Three scoped variants of `select` exist: `select_all`, `select_if`, and `select_at`

**Notes**

Special regex-like functions exist to grab specific data frame columns by name or pattern
```
data %>% select(contains("."))
data %>% select(everything())
data %>% select(starts_with("abc"))
data %>% select(num_range("x", 1:5))
data %>% select(ends_with("xyz"))
data %>% select(matches(".t."))
data %>% select(one_of(c("Sp", "Gen")))
```

# Filter

**Objective**: Reduce data frame observations according to a specific value or condition

**Function**: 
```
filter(data, ...)
```

**Arguments**:
```
data:           data frame
...:            conditions to be met
```

**Examples**:

```{r}
airquality.filtered <- as_tibble(airquality) %>% 
  filter(Ozone > 30, Solar.R > 200)
```

**Useful logical and comparison operators**

```
<   Less than                    !=      Not equal to
>   Greater than                 %in%    Group membership
==  Equal to                     is.na   is NA
<=  Less than or equal to        !is.na  is not NA
>=  Greater than or equal to     &,|,!   Boolean operators
```

operations exist for numeric vector comparison as well
```
near(x, y, tol)
between(x, left, right)
```

# Group By

**Objective**: group observations based on unique values of a categorical feature (or features).

**Function**: 
```
group_by(data, ...)
```

**Arguments**:
```
data:           data frame
...:            categorical variable(s) to group by
```

**Examples**:

```{r}
library(ggplot2)
```

summarize the data by groups. One row for each group, and one column for each function applied.
```{r}
diamonds %>% 
  group_by(cut) %>% 
  summarize(carat = mean(carat), 
            depth = mean(depth),
            price = mean(price))
```

add summarization information back onto the original data frame
```{r}
mtcars %>% 
  group_by(cyl) %>% 
  mutate(mean_mpg = mean(mpg), sd_mpg = sd(mpg))
```

# Summarise

**Objective**: produce summary statistics for the data

**Function**: 
```
summarise(data, ...)
```

**Arguments**:
```
data:           data frame
...:            name-value pairs of summarize functions. The name is name of the resulting variable, and the value is a single value that results from the application of a summarization function to the data
```

**Examples**:

summarize over all examples in the data frame
```{r}
by_cyl <- mtcars %>% 
  summarize(mean_mpg = mean(mpg, na.rm = TRUE),
            sd_mpg = sd(mpg, na.rm = TRUE))
```

summarize using grouped examples in the data frame
```{r}
by_cyl <-mtcars %>% 
  group_by(cyl) %>% 
  summarize(mean_mpg = mean(mpg, na.rm = TRUE),
            sd_mpg = sd(mpg, na.rm = TRUE))
```

**Scoped Variants**

- `summarise_all()` will summarise all columns based on your further instructions
- `summarise_if()` requires a function that returns a boolean. If that is true, the summary instructions will be followed.
- `sumarise_at()` requires you to specify columns inside a `vars()` argument for which the summary will be done.

**Useful Summarization Functions**:
```
Center: mean(), median()
Spread: sd(), IQR(), mad()
Range: min(), max(), quantile()
Position: first(), last(), nth(),
Count: n(), n_distinct()
Logical: any(), all()
```

# Arrange

**Objective**: sort data by row(s) in ascending or descending order

**Function**: 
```
arrange(data, ...)
```

**Arguments**:
```
data:           data frame
...:            variable(s) to order by
```

**Examples**:

in ascending order
```{r}
by_mpg <- mtcars %>% 
  arrange(mpg)
```

in descending order
```{r}
by_mpg <- mtcars %>% 
  arrange(desc(mpg))
```

**Notes**:

If you already grouped your data, you can refer to that group within the `arrange()` statement as well by adding a `.by_group = TRUE` statement

# Pull

**Objective**: Return a single column from a data frame as a vector (and not as a dataframe like with `select()`)

**Function**:
```
pull(.data, var = -1)
```

**Arguments**:
```
data:           data frame
var:            a literal variable name, a positive integer (from left), or a negative integer (from right)
```

**Examples**

pull by variable name
```{r, echo=TRUE, results="hide"}
mtcars %>% 
  pull(cyl)
```

or by variable index
```{r, echo=TRUE, results="hide"}
mtcars %>% 
  pull(1)
```

The analogpus base R operation is
```{r, echo=TRUE, results="hide"}
mtcars[["cyl"]]
```

or by index
```{r, echo=TRUE, results="hide"}
mtcars[[1]]
```

# n_distinct

**Objective**: find the number of distinct (or unique) elements of a list or vector, or data frame column

**Function**: 
```
n_distinct(..., na.rm = FALSE)
```

**Arguments**:
```
...           vectors of values
na.rm         if TRUE missing values don't count
```

**Example**

```{r}
mtcars$mpg %>% n_distinct()
```

possible to achieve the sane count in base R
```{r}
mtcars$mpg %>% unique %>% length
```

# Mutate
- create features from existing features

# Transmute
- like mutate, but drops the existing variables in favor of newly defined ones.

# Rename
- rename an existing variable

# Slice
- select specifc rows from data frame

# sample_frac
- randomly select a fraction of rows of data

# sample_n
- randomly select n rows of data

# Join
- inner, outer, left, right
