---
title: "ggplot2 Notes"
author: "Christopher Willis"
output:
  html_document:
    theme: united
    highlight: tango
    df_print: paged
    toc: true
    toc_float: true
    collapsed: false
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = FALSE)
```

# The ggplot2 Package

```{r, echo=TRUE, results="hide"}
library(ggplot2)
```

# Aesthetics

An aesthetic is a visual property of the objects in your plot; e.g. displays points are circles or triangles

possible to map a variable to color aesthetic
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class))
```

... or size aesthetic. Also possible to set `alpha` (transparancy) or `shape` (type of point)
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class))
```

... also possible to set the aesthetic properties of the geom manually
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```

# Facets

possible to split a plot into factets; i.e. one subplot subsetted for each categorical variable
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)
```

# Geometric Objects

Every geom_* function in ggplot2 takes a mapping argument

possible to fit a smooth line to data with geom_smooth()
```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

possible to separate variable into different lines based on a categorical varible
```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv))
```

# Statistical Transformations

- bar charts, histograms, polygons bin the data and plot bin counts   
- smoothers fit model to data and plot predictions   

- "stat" algorithm is used in each case; i.e. transformation applied to data to compute summary statistics   
- geom uses transformed data to build plot   
- possible to change default stat for a given geom   

## Bar Charts

bar chart calculates counts from the dataset automatically
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut))
```

## Histograms

use bar charts for discrete data, histograms for continuous
```{r}
ggplot(diamonds, aes(price, fill = cut)) +
  geom_histogram(binwidth = 500)
```

## Boxplots

boxplots can be used to compute summary statistics

## Summary

```{r}
ggplot(data = diamonds) + 
  stat_summary(
    mapping = aes(x = cut, y = depth),
    fun.ymin = min,
    fun.ymax = max,
    fun.y = median
  )
```

# General ggplot2 format

Any ggplot2 plot takes 7 arguments. The template is

```
ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(
     mapping = aes(<MAPPINGS>),
     stat = <STAT>, 
     position = <POSITION>
  ) +
  <COORDINATE_FUNCTION> +
  <FACET_FUNCTION>
```

# using factors

```{r}
library(dplyr)
air.temp <- airquality %>%  
  as_tibble %>% 
  mutate(tempcat = factor(Temp > 80, labels = c("cold", "hot")),
         Year = 1973,
         date = ISOdate(Year, Month, Day))
```

```{r}
air.temp %>% 
  ggplot() + 
  geom_point(mapping = aes(x = date, y = Temp, color = tempcat))
```

