---
title: "tidyr Notes"
author: "Christopher Willis"
output:
  html_document:
    theme: united
    highlight: tango
    df_print: paged
    toc: true
    toc_float: true
    collapsed: false
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = FALSE)
```

# The tidyr Package

```{r, echo=TRUE, results="hide"}
library(tidyr)
```

The `tidyr` package is built for the sole purpose of simplifying the process of creating tidy data   

`tidyr` provides four fundamental functions for data tidying:   
- `gather()` makes "wide" data longer   
- `spread()` makes "long" data wider   
- `separate()` splits a single column into multiple columns   
- `unite()` combines multiple columns into a single column

`gather` and `spread` are used to resolve one of two common problems:   
- One variable might be spread across multiple columns.   
- One observation might be scattered across multiple rows.   

# Gather

**Objective**: Sometimes data is formatted in such a way that some of the columns are named after values of variable takes on instead of the variable itself (e.g. columns for whether the year is `1999`, `2000`, etc. where the desired variable is `date` in years). Here, `gather` can be used to gather those columns into a new pair of variables.

**Function**: 
```
gather(data, key, value, ..., na.rm = FALSE, convert = FALSE)
```

**Arguments**:
```
data:           data frame  
key:            column name representing new variable   
value:          column name representing variable values  
...:            names of columns to gather (or not gather). If empty, all columns are selected. Use a:b to select range or -c to exclude.  
na.rm:          option to remove observations with missing values  
convert:        if TRUE will automatically convert values to logical, integer, numeric, complex or factor as appropriate
```

**Example**:

<!-- One variable might be spread across multiple columns. -->

<!-- one variable (e.g. date) might be spread across multiple columns (e.g. 1999, 2000, etc.). Use gather to pull the multiple columns into a single column. -->

The `gather` operation requires at least three arguments.   
- The set of columns that represent values and not variables; i.e. `X:Z`   
- The key, or the name of the variable whose values form the column names; i.e. `stock`   
- The value, or the name of the variable whose values are spread over the cells; i.e. `price`   

Here, the `price` of each stock is spread across three columns
```{r}
stocks <- tibble(time = as.Date('2009-01-01') + 0:9,
                 X = rnorm(10, 0, 1),
                 Y = rnorm(10, 0, 2),
                 Z = rnorm(10, 0, 4))
```

The resulting dataframe is
```{r}
stocks
```

Using `gather`, we can pull each stock into a single stock column
```{r}
stocks %>% gather(key = "stock", value = "price", X:Z)
```

**Notes**:

These all produce the same result
```
stocks %>% gather("stock", "price", X:Z)
stocks %>% gather("stock", "price", -time)
stocks %>% gather("stock", "price", X, Y, Z)
stocks %>% gather("stock", "price", 2:4)
```

Other names for gather include melt (reshape2), pivot (spreadsheets), and fold (databases)

# Separate

**Objective**: Split a single variable into two; i.e. turn a single character column into multiple columns.

**Function**: 
```
separate(data, col, into, sep = " ", remove = TRUE, convert = FALSE)
```

**Arguments**:
```
data:           data frame
col:            column name or position representing current variable
into:           names of new variables to create as character vector
sep:            how to separate current variable (char, num, or symbol); default is a regular expression that matches any sequence of non-alphanumeric values
remove:         if TRUE, remove input column from output data frame
convert:        if TRUE will automatically convert values to logical, integer, numeric, complex or factor as appropriate
```

**Example**:

```{r}
revenue <- tibble(Group = c(1, 1, 1, 1, 2, 2),
                  Year = c(2006, 2007, 2008, 2009, 2006, 2007),
                  Quarter = c("Qtr.1", "Qtr.1", "Qtr.1", "Qtr.1", "Qtr.1", "Qtr.1"),
                  Revenue = c(15, 12, 22, 10, 12, 16))
revenue
```

```{r}
revenue %>% 
  separate(col = Quarter, 
           into = c("Interval", "Interval_ID"), 
           sep = "\\.", 
           convert = TRUE)
```

# Unite

**Objective**: Merge two variables into one

**Function**: 
```
unite(data, col, ..., sep = " ", remove = TRUE)
```

**Arguments**:
```
data:           data frame
col:            column name of new "merged" column
...:            names or positions of columns to merge
sep:            separator to use between merged values
remove:         if TRUE, remove input column from output data frame
```

**Example**:

```{r}
revenue <- tibble(Group = c(1, 1, 1, 1, 2, 2),
                  Year = c(2006, 2007, 2008, 2009, 2006, 2007),
                  Interval = c("Qtr", "Qtr", "Qtr", "Qtr", "Qtr", "Qtr"),
                  Interval_ID = c(1, 1, 1, 1, 1, 1),
                  Revenue = c(15, 12, 22, 10, 12, 16))
revenue
```

```{r}
revenue %>% unite(col = Quarter, 
                  c(Interval, Interval_ID), 
                  sep = ".")
```

**Notes**:

This function is a complement to `separate()`

# Spread

**Objective**: Reshape long format data to wide format data; opposite of gather.

**Function**: 
```
spread(data, key, value, fill = NA, convert = FALSE)
```

**Arguments**:
```
data:           data frame
key:            column values to convert to multiple columns
value:          single column values to convert to multiple columns' values 
fill:           If there isn't a value for every combination of the other variables and the key column, this value will be substituted
convert:        if TRUE will automatically convert values to logical, integer, numeric, complex or factor as appropriate
```

**Example**:

```{r}
measurements <- tibble(name = c("Wilbur", "Petunia", "Gregory", "Wilbur", "Petunia", "Gregory"),
                       drug = c("a", "a", "a", "b", "b", "b"),
                       heartrate = c(67, 80, 64, 56, 90, 50))
```

```{r}
measurements %>% spread(key = drug, value = heartrate)
```

```{r}
measurements %>% spread(key = name, value = heartrate)
```

**Notes**:

This function is a complement to `gather()`

# Drop NA

**Objective**: drop data frame rows with missing values

**Function**: 
```
drop_na(data, ...) 
```

**Arguments**: 
```
data          A data frame.
...           A selection of columns. If empty, all variables are selected. 
```

**Example**:

Drop `NA` values appearing in the `Ozone` variable
```{r}
airquality %>% drop_na(Ozone)
```

Drop `NA` values appearing in all variables
```{r}
airquality %>% drop_na
```
